---
title: "Species richness trends for Bavaria"
author: "RS-eco"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Species richness trends for Bavaria}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = T, comment = NA, warning=F, message=F, echo=T, error=F, eval=F,
  comment = "#>", fig.width=7, fig.height=7, fig.path="../figures/"
)
```

# Data Analysis

Here, I perform a first analysis on species richness changes across Bavaria using observations from the ASK database.

## Load data

First, we load and subset the data:

```{r load_data}
rm(list=ls()); gc()

#+ data_setup, results="hide"
# Load packages
library(data.table); library(dtplyr); library(dplyr, warn.conflicts = FALSE)
library(magrittr); library(tidyr)

########################

# Load ASK database
my_db <- DBI::dbConnect(RSQLite::SQLite(), dbname = "../rawdata/ASK.db")
# Pull part of ASK database including data on species 
ask_data <- dplyr::tbl(my_db, paste("ask_art")) %>% collect()

# Load background grid
tk25 <- dplyr::tbl(my_db, "geo_tk25_quadranten") %>% dplyr::collect() %>%
  tidyr::unite(col="quadrant", c("KARTE", "QUADRANT"), sep="/", remove = TRUE) %>% 
  dplyr::select(-c("KARTE_QUAD")) 

# Load taxonomy
library(bavDC)
data("taxonomyStd")

# Load ask_fuo
ask_fuo <- dplyr::tbl(my_db, "ask_fuo") %>% collect()
DBI::dbDisconnect(my_db)

# combine gridded map from dat_ask with full locations of recorded species and taxonomy
dat_ask <- ask_data %>% dplyr::select(-c(gkk_rw, gkk_hw, objnr, rw, hw)) %>% 
   dplyr::left_join(ask_fuo) %>%  dplyr::left_join(tk25) %>%
   dplyr::left_join(taxonomyStd, by=c("art" = "scientificName"))
dat_ask <- dat_ask %>% tidyr::drop_na(class, order)
dat_ask$KARTE_QUAD <- as.numeric(sub("/", "", dat_ask$quadrant))

#nrow(dat_ask)
ask_tk25 <- dat_ask %>% dplyr::select(KARTE_QUAD) %>% drop_na()
#nrow(ask_tk25)

ask_gkk <- ask_data %>% dplyr::select(gkk_rw, gkk_hw) %>% drop_na()
#nrow(ask_gkk)

# Create custom taxon vector (Aves, Lepidoptera, Odonata, Orthoptera)
dat_ask$class_order <- "Aves"
dat_ask$class_order[dat_ask$class == "Insecta"] <- dat_ask$order[dat_ask$class == "Insecta"]
#unique(dat_ask$class_order)

dat_ask %<>% dplyr::select(XQMITTE, YQMITTE, jahr, mon, class_order, scientificNameStd, art)
rm(ask_fuo, ask_data, my_db); invisible(gc())

# ## Subset data by year
dat_ask <- dat_ask %>% filter(jahr >= 1985, jahr <= 2019); invisible(gc())
```

## Calculate richness over time

Split the data into the different taxonomic groups, calculate the species richness over time and turn them into a raster stack:

```{r}
# Calculate species richness for each taxonomic group
sum_dat1 <- dat_ask %>% 
  group_by(XQMITTE, YQMITTE, jahr, class_order) %>% 
  summarise(sr_std=n_distinct(scientificNameStd)) %>% drop_na() %>% ungroup() %>%
  arrange(jahr) %>% pivot_wider(names_from=jahr, values_from=sr_std) %>%
  mutate_at(vars(-c(XQMITTE, YQMITTE, class_order)), ~replace_na(., replace = 0))

list_dat1 <- sum_dat1 %>% group_by(class_order) %>% group_split(.keep=F)
r_names1 <- sum_dat1 %>% group_by(class_order) %>% group_keys() %>% unlist()
rast_dat1 <- lapply(list_dat1, function(x){
  dat <- raster::rasterFromXYZ(x, crs=sp::CRS("+init=epsg:31468"), res=c(6100, 5550), digits=0)
  raster::extend(dat, y=raster::extent(c(4279033, 4638933, 5236691, 5608687)))
})
names(rast_dat1) <- r_names1

# Calculate total species richness
sum_dat2 <- dat_ask %>% 
  group_by(XQMITTE, YQMITTE, jahr) %>% 
  summarise(sr_std=n_distinct(scientificNameStd)) %>% drop_na() %>% ungroup() %>%
  arrange(jahr) %>% pivot_wider(names_from=jahr, values_from=sr_std) %>%
  mutate_at(vars(-c(XQMITTE, YQMITTE)), ~replace_na(., replace = 0))
rast_dat2 <- raster::rasterFromXYZ(sum_dat2, crs=sp::CRS("+init=epsg:31468"), 
                                   res=c(6100, 5550), digits=0)
rast_dat2 <- raster::extend(rast_dat2, y=raster::extent(c(4279033, 4638933, 5236691, 5608687)))
rast_dat <- c(rast_dat1, list(rast_dat2))
names(rast_dat)
names(rast_dat)[5] <- "Total"
```

## Perform linear regression across raster stack

<!-- Weight by sampling effort -->

```{r}
library(raster)
### A much (> 100 times) faster approach is to directly use 
### linear algebra and pre-compute some constants
time <- 1:raster::nlayers(rast_dat[[1]])

## add 1 for a model with an intercept
X <- cbind(1, time)

## pre-computing constant part of least squares
invXtX <- solve(t(X) %*% X) %*% t(X)

## much reduced regression model; [2] is to get the slope
slope <- function(y) (invXtX %*% y)[2]
intercept <- function(y) (invXtX %*% y)[1]

slope_lm_dat <- raster::stack(lapply(1:length(rast_dat), function(z){
  raster::calc(rast_dat[[z]], slope)
}))
names(slope_lm_dat) <- names(rast_dat)
intercept_lm_dat <- raster::stack(lapply(1:length(rast_dat), function(z){
  raster::calc(rast_dat[[z]], intercept)
}))
names(intercept_lm_dat) <- names(rast_dat)
```

## Plot Slope and intercept

### Individual taxa

```{r, fig.width=7, fig.height=10}
par(mfrow=c(4,2))
for(z in 1:4){
  plot(intercept_lm_dat[[z]], main=paste0("Intercept - ", names(rast_dat)[[z]]))
  plot(slope_lm_dat[[z]], main=paste0("Slope - ", names(rast_dat)[[z]]))
}
```

### Total species richness

```{r, fig.width=7, fig.height=5}
par(mfrow=c(1,2))
plot(intercept_lm_dat[[5]], nc=1, main="Intercept")
plot(slope_lm_dat[[5]], nc=1, main="Slope")
rm(list=ls()); gc()
```
